R Notebook

### 2025-03-05 Scripts/authoritarian_only_latex_2.Rmd
Good clarifications to theory section; end of major post CPS revisions.

Along with the political benefits it may provide, election manipulation is generally characterized as having both operational costs (which the incumbent may be unwilling or unable to pay [@Greene2010]) and legitimacy costs which reduce other actors' level of cooperation with the incumbent [@birch2011]. The most visible, and arguably most studied, form of such legitimacy costs is mass post-election protest against the incumbent. Here, I follow the broad consensus in the literature that authoritarian incumbents find protest undesirable, since they entail some risk of incumbent defeat and may require costly repression or concessions. I assume, as a result, that incumbents seek to demobilize protest; protest movements that survive longer indicate the incumbent is relatively weaker compared to the opposition. By contrast, when protest movements meet a quick demise, the incumbent can be considered relatively strong.^[In both cases, this assume that incumbent survives the protest movement, by far the modal outcome. When the opposition wins, an outcome considered in some of the empirical models below, the assumption is reversed: a quicker opposition victory indicates a weaker incumbent.] How might election manipulation be connected to the incumbent's ability to quickly defeat a protest movement?

Prior models of electoral protest by strategic opposition actors, discussed above, treat election manipulation and genuine votes as substitute goods. For any given vote-share, an increase in manufactured votes reduces the number of genuine votes. For classic models, which consider manufactured votes to be merely illusions of strength, the higher the number of manufactured votes for any given vote total, the weaker the incumbent. This means that strategic opposition actors are more likely to initiate protest when they perceive the incumbent to be weaker, and expect to be more likely to succeed in a challenge.

In the classic model, the incumbent's (unknown) true popularity is positively related to its vote-share in the election, and negatively related to its use of fraud. This dynamic is the fulcrum upon which these models hinge: less popular incumbents (who are expected to be less likely to defeat an opposition protest) must manipulate more in order to win elections. Popular incumbents need manipulate less. Opposition actors gauge the probability of an pro-government protest outcome by estimating the incumbent's true popularity based on the degree of manipulation and the incumbent's vote-share in the election. A high vote-share with low fraud indicates a more popular incumbent, likely deterring protest. A low vote-share with a high degree of fraud indicates a very weak incumbent, making strategic protest more likely. In other words, manipulated votes are empty in the classic model; subtracting them from the incumbent's vote-total gives opposition actors their estimate of the incumbent's ability to win a post-election standoff.

Thus the predicted relationship between manipulation and protest likelihood in the classic model is positive: more manipulation makes protest more likely. However, as increasing vote-share increases the number of genuine votes for the incumbent at any fixed level of fraud, the effect of fraud should be conditional on the claimed vote-share. As Rozenas [@Rozenas2016b] notes, if an incumbent wins an election with low manipulation, it generates a strong signal of its popularity; if it wins a heavily manipulated election, this signal is weaker. As an illustration, if the incumbent's total vote-share is 70 percent, for example, of which the opposition estimates 15 percent to be manufactured, the incumbent's genuine popularity is estimated at a respectable 55 percent. By contrast, if the incumbent posts a 50 percent margin of victory with the same 15 percent estimated to be manufactured, the opposition may perceive that the incumbent is weak in its genuine support (35 percent). This produces a first, baseline hypothesis, derived from the classic model.

@@

It is important to note that the same hypothesis is also generated by non-centralized, non-strategic models of election protest. The mechanism, however, is different. Here, the combination of low incumbent vote-share and high manipulation is indicative of widespread grievances coordinated a 'stolen' election. These are conditions that should be most likely to produce protest [@tucker2007]. By contrast, as vote-share increases for a set level of fraud, the incumbent's genuine popularity is higher and thus grievances are lower (reducing the likelihood of protest).

Integrating Simpser's argument with recent research on the principal-agent dynamics inherent to election manipulation provides different insights. Where Simpser argues that manipulation should be blatant and excessive (that is, producing a large margin of victory) to send signals of strength, the principal-agent framework for understanding manipulation suggests otherwise. This research trend emphasizes the degree to which political principals require the cooperation of large numbers of front-line agents in order to generate extensive election fraud, agents who may shirk if their perceived interests are not aligned with the incumbent [@rundlett2016deliver]. Factors that minimize principal-agent problems have been shown to facilitate greater election manipulation. These include the incumbent's local popularity [@rundlett2016deliver], the incumbent's ability to distribute patronage resources [@Harvey2019;@hale_patronal_2014], and low risk of punishment for pro-regime lawbreaking [@Harvey2022]. As discussed below, these are also independently useful for suppressing post-election protest. Consequently, if an incumbent is able to manufacture a large numbers of votes, it is also likely to have the resource capacity, control over bureaucratic institutions, and perhaps deep wells of popularity in parts of the country that can serve as independent sources of regime durability, even if the incumbent's nationwide margin of victory is low.

The initial emergence of protest triggers a response by the government and ruling party that can take the form of repression, concessions, or toleration [@tillyMobilizationRevolution1978;@lipskyProtest1968;@Hummel2019;@raslerConcessionsRepression1996;@Bishara2015;@Frantz2014]. Concessions may be aimed at protesting parties themselves [@reuter2015;@mooreRepressionDissentSubstitution2000;@Hafner-Burton2016;@Piven1979] or the more moderate public, in order to keep them on the sidelines of the conflict [@Hummel2019]. In either case, large-scale election manipulation indicates control over considerable patronage resources [@Harvey2019;@hale_patronal_2014;@greene_why_2007], which in turn indicates the capacity to engage in such payments (or withhold them from clients in exchange for passive support [@Rosenfeld2017]).

Ruling parties may also lean on repression, calling upon the security services to disperse protests, arrest leaders, and perhaps employ harsher methods including disappearance, torture, and execution [@Escriba_Folch2013;@Davenport2007]. Though such crackdowns risk provoking a backlash, they are aimed at demobilizing the opposition by increasing the costs of participation in overt dissent [@DeVogel2022]. The ability to protect regime agents from any legal consequences for carrying out orders, a condition which helps enable illegal election manipulation [@Harvey2022], is also an asset when asking law enforcement agents to carry out possibly violent repression [@MAGALONI_RODRIGUEZ_2020;@stavro_police_2023;@abouharbFactoJudicialIndependence2013].

@@

This argument\textemdash that successful, large-scale election manipulation draws on the same resources a state may use to suppress protest\textemdash turns the deterrent logic on its head. Instead of signalling weakness, large-scale fraud indicates a well of resilience: control over resources, impunity for pro-regime lawbreaking, and loyal agents. As a result, governments that build and deploy an electoral manipulation apparatus are simultaneously innoculating themselves against the cost of protest.

If election manipulation is an indicator of underlying capacities the incumbent may use to defeat protest movements, more severe electoral manipulation will be associated with shorter-lived protest waves. Manufactured votes are not negatively associated with genuine popularity, but positively associated with impunity for agents and control over patronage resources. They are also positively associated with incumbent popularity [@rundlett2016deliver]. Abundant resources and extensive impunity reduce the principal-agent problems that frustrate election manipulation efforts. These same resources, institutions, and loyal agents make protest outcomes less likely to be favorable for the opposition. Note that, in this model, protest initiation is not tied to expected protest outcome; protest is driven by non-strategic concerns related to the incumbent's popularity (i.e. citizens' diverse grievances) and whether manipulation is perceived as having been decisive (i.e. a stolen election). Incumbents with the resources and institutions needed to produce widespread election manipulation can draw on those same capacities to buy off, repress, or otherwise quash post-election protest. This yields Hypothesis 2.

@@

The classic model holds that manipulated votes correlate negatively with incumbents' post-election strength, implying that when incumbents rely heavily on election manipulation they will struggle to put down post-election protests. This effect should likewise be conditional on incumbent vote-share, since those incumbents who produce small margins of victory while relying heavily on manipulation will have limited ability to push back successfully against protest movements. A higher likelihood of pro-government protest outcome reduces the likelihood that the opposition will initiate protest. Consequently, classic models predict that more election manipulation should be associated with longer-lived protest movements against surviving incumbents.^[In the appendix, I check the separate prediction that manipulation should be positively associated with incumbents being ousted by protest movements.]

The argument can be summarized as follows. Incumbents will generally seek to manipulate elections as much as they can, to capitalize on both the direct and indirect benefits of electoral manipulation [@simpser2013]. To do so, they must solve principal-agent problems by consolidating control over patronage resources and shielding agents from negative consequences of their actions. The same resources that can be used to manipulate elections on a wide scale\textemdash including natural popularity, resource control, and control over the legal apparatus\textemdash also better enable incumbents to dismantle protest movements when they do emerge. As a result, more extensive manipulation should be associated with shorter-lived protest movements against incumbents. Much election-related protest, then, is unrelated to strategic concerns about the likelihood of protest success.

>>>>

Election manipulation is generally characterized as having legitimacy costs which reduce other actors' level of cooperation with the incumbent [@birch2011]. The most visible, and arguably most studied, form of such legitimacy costs is mass post-election protest against the incumbent. Here, I do not challenge the broad consensus in the literature that authoritarian incumbents find sustained, sizable protests undesirable, since they entail some risk of incumbent defeat and may require costly repression or concessions. I assume, as a result, that incumbents generally seek to demobilize protest; protest movements that survive longer indicate the incumbent is relatively weaker compared to the opposition.^[The implications of protest toleration by incumbents will be discussed later.] By contrast, when protest movements meet a quick demise, the incumbent can be considered relatively strong.^[In both cases, this assume that incumbent survives the protest movement, by far the modal outcome. When the opposition wins, an outcome considered in some of the empirical models below, the assumption is reversed: a quicker opposition victory indicates a weaker incumbent.] As argued below, electoral manipulation can be demonstrative of capacities that may also be used to defeat protest movements.

Regime-weakness models of electoral manipulation, discussed above, treat election manipulation and genuine votes as substitute goods. For any given vote-share, an increase in manufactured votes reduces the number of genuine votes. In these models, which consider manufactured votes to be merely illusions of strength, the higher the number of manufactured votes for any given vote total, the weaker the incumbent. This means that strategic opposition actors are more likely to initiate protest when they perceive the incumbent to be weaker, and expect to be more likely to succeed in a challenge. A high vote-share with low fraud indicates a more popular incumbent, likely deterring protest. A low vote-share with a high degree of fraud indicates a very weak incumbent, making strategic protest more likely. In other words, manipulated votes are empty in the classic model; subtracting them from the incumbent's vote-total gives opposition actors their estimate of the incumbent's ability to win a post-election standoff.

Thus the predicted relationship between manipulation and protest likelihood in the regime-weakness model is positive: more manipulation makes protest more likely for a given vote-share. However, the inverse is also to be expected: as incumbent vote-share increases for any fixed level of fraud, the opposition should perceive a stronger incumbent and be less likely to protest. In sum, the effect of fraud should be conditional on the claimed vote-share [@Rozenas2016b]. This produces a first, baseline hypothesis.

@@

It is important to note that the same hypothesis is also generated if we assume that protest initiation is broadly, rather than narrowly, strategic. The mechanism, however, is different. Here, the argument is that widespread grievances result in low natural popularity for the incumbent; at election time, the incumbent must compensate for this low popularity by manipulating the election. Protest initiation is made more likely both by the population's grievances alone [@Beaulieu2014a;@Brancati2016], and by the coordination device provided by the 'stolen' election [@tucker2007]. As natural vote-share increases for a fixed level of manipulation, the likelihood of protest initiation should decline due to reduced grievances; for a fixed level of popularity, increasing manipulation should increase the likelihood of protest by raising the plausibility of a stolen election among dissatisfied citizens.

We should only expect H1 to be falsified in the event that protest is driven by narrowly strategic concerns *and* election manipulation is indicative of regime strength. In this case, opposition actors should be deterred from launching protest movements as increasing manipulation indicates deeper regime resilience (in the form of resources, weak rule of law, loyalty of agents, and so on). The regime-strength model, as I call it, is defended in more detail below.

## Developing the regime-strength model of electoral manipulation

Simpser's foundational argument in the regime-strength model holds that manipulated elections can send costly signals of the incumbent's "capacities, resources, [and] dispositions" that will help it retain power in the future [@simpser2013, p. 80]\textemdash that is, of its political strength. Additional research focusing on the principal-agent dyanmics inherent in election manipulation reinforces the idea that manipulation carries such costly signals, such that more extensive electoral manipulation indicates greater regime strength. This research trend emphasizes the degree to which political principals require the cooperation of large numbers of front-line agents in order to generate extensive election fraud, agents who may shirk if their perceived interests are not aligned with the incumbent [@rundlett2016deliver].

Electoral manipulation has been shown to be more extensive under conditions that minimize that minimize principal-agent problems (i.e. those that make low-level agents more likely to cooperate with the incumbent). These include: greater local popularity for the incumbent [@rundlett2016deliver], greater incumbent control over patronage resources [@Harvey2019;@hale_patronal_2014], and lower risk of punishment for pro-regime lawbreaking [@Harvey2022]. As discussed below, these capacities and resources are also independently useful for surviving post-election protest. Consequently, if an incumbent is able to manufacture a large numbers of votes, it is also likely to have the resource capacity, control over bureaucratic institutions, and perhaps deep wells of popularity in parts of the country that can serve as independent sources of regime durability, even if the incumbent's nationwide margin of victory is low. Lastly, extensive manipulation also indicates that large number of agents assess that the incumbent will remain in power in the future [@simpser2013], which itself becomes a self-reinforcing equilibrium [@hale_patronal_2014].

These resources and capacities improve the incumbent's ability to respond to a nascent protest movement through repression or concessions.^[Toleration, another possibility, is discussed later.] [@tillyMobilizationRevolution1978;@lipskyProtest1968;@Hummel2019;@raslerConcessionsRepression1996;@Bishara2015;@Frantz2014]. Concessions may be aimed at protesting parties themselves [@reuter2015;@mooreRepressionDissentSubstitution2000;@Hafner-Burton2016;@Piven1979] or the more moderate public, in order to keep them on the sidelines of the conflict [@Hummel2019]. In either case, large-scale election manipulation indicates control over considerable patronage resources [@Harvey2019;@hale_patronal_2014;@greene_why_2007], which in turn indicates the capacity to engage in such payments (or withhold them from clients in exchange for passive support [@Rosenfeld2017]).

Ruling parties may also lean on repression, calling upon the security services to disperse protests, arrest leaders, and perhaps employ harsher methods including disappearance, torture, and execution [@Escriba_Folch2013;@Davenport2007]. Though such crackdowns risk provoking a backlash, they are aimed at demobilizing the opposition by increasing the costs of participation in overt dissent [@DeVogel2022]. The ability to protect regime agents from any legal consequences for carrying out orders, a condition which helps enable illegal election manipulation [@Harvey2022], is also an asset when asking law enforcement agents to carry out possibly violent repression [@MAGALONI_RODRIGUEZ_2020;@stavro_police_2023;@abouharbFactoJudicialIndependence2013]. Additionally, more consolidated control over patronage resources can help incentivize elite unity [@Reuter2019;@hale_patronal_2014;@Arriola_2009], which can enable repression by the security forces [@hendrix_2017;@Frugé_2019;@Albrecht_Ohl_2016].

@@

This argument\textemdash that successful, large-scale election manipulation draws on the same resources a state may use to suppress protest\textemdash turns the deterrent logic on its head. Instead of signalling weakness, large-scale fraud indicates a well of resilience: control over resources, impunity for pro-regime lawbreaking, and loyal agents. As a result, governments that build and deploy an electoral manipulation apparatus are simultaneously inoculating themselves against the expected cost of protest.

Putting it all together: this theory assumes that authoritarian incumbents generally prefer order to unrest, and that protesters\textemdash who under either a narrow or broadly strategic view are aiming for some kind of concession from the incumbent\textemdash want to demonstrate the popularity of their cause through large, sustained protests. Electoral protests are a form of competition between the incumbent and opposition actors, in which both sides are trying to demonstrate their strength in order to win concessions from\textemdash or defeat outright\textemdash the other [@Beaulieu2014a]. The outcome of this contest reveals the underlying strengths of the participants. Consequently, governments that quickly defeat protest movements can be considered stronger than those that take weeks or months to prevail, which in turn can be considered stronger than those governments which collapse in the face of mass protest.  Incumbents with the resources and capacities needed to produce widespread election manipulation can draw on those same tools to buy off, repress, or otherwise quash post-election protest. This yields Hypothesis 2.

@@

The argument can be summarized as follows. Incumbents will generally seek to manipulate elections as much as they can, to capitalize on both the direct and indirect benefits of electoral manipulation [@simpser2013]. To do so, they must solve principal-agent problems by consolidating control over patronage resources and shielding agents from negative consequences of their actions. The same resources that can be used to manipulate elections on a wide scale\textemdash including natural popularity, resource control, and control over the legal apparatus\textemdash also better enable incumbents to dismantle protest movements when they do emerge. As a result, more extensive manipulation should be associated with shorter-lived protest movements against incumbents.^[In the appendix, I check the separate prediction that manipulation should be positively associated with incumbents being ousted by protest movements.]

One challenge to this theory is the prospect that incumbents may choose to tolerate protests, rather than attempting to repress them or co-opt their participants. This concern should be modest for the cases being considered here. First, toleration is most likely when protests impose few costs on the regime and make only targeted demands [@kleinDynamicsPoliticalProtests2018]. National elections are highly politically salient, and challenging their validity gets to the core of electoral authoritarian incumbent's claim to authority. As such, tolerating or ignoring protest is unlikely in this setting. In such cases, we should expect to see apparent 'toleration' displaced by 'attrition'\textemdash efforts by the incumbent to increase the cost of protest while avoiding open repression [@yuenNeitherRepressionConcession2017]. In other words, even in the absence of explicit repression or concessions, we should still expect incumbents to attempt to defuse protest movements aimed at challenging the most recent election.

Testing H1 and H2, together, will help to evaluate the regime-weakness model of electoral manipulation. If H1 is supported and H2 is not, it would provide evidence in favor of the regime-weakness model; manipulation would appear associated with incumbent weakness, and protest initiation would appear to take advantage of such. On the other hand, if both hypotheses are supported, it would endorse the argument that electoral manipulation can be associated with sources of incumbent strength in a world where the aims of electoral protest are broadly (rather than narrowly) strategic.
